R Notebook

### 2025-03-03 Scripts/authoritarian_only_latex_2.Rmd
Edits to lit review: deterrence logic based on 2 assumptions

# Literature review: three views of election manipulation and mass protest

In classic models of election manipulation, manipulation is a short-term tool used instrumentally to secure victory in the election [@lehoucq2002]. Much of the work in this tradition has been elaborated through formal models. Though there are, of course, many variations, in several such models incumbents are characterized as 'weak' or 'strong' based on their true support in the electorate. This true level of support can be obscured prior to elections due to pervasive preference falsification [@Kuran1991] and the low level of political information common in authoritarian regimes [@guriev2020]. However, low popularity (i.e. weakness, in these models) can be revealed when incumbents need to resort to fraud in order to secure re-election. As a result, the opposition perceives election fraud as a reliable signal of incumbent weakness, and will stage protests in order to push out the weak incumbent. For example, the model developed by Little, Tucker, and LaGatta [-@Little2015] predicts that "Just as stronger election results for the incumbent lead to less protest because they indicate that she is popular, reports of more fraud lead to more protest because they indicate that the incumbent is, for a given election result, less popular" (p. 1153). In equilibrium, this can lead weaker incumbents to step down rather than face protest due to massive fraud\textemdash a pro-democracy outcome. Similar results are developed by Chernykh and Svolik [-@chernykh2015third], Luo and Rozenas [-@Luo2020] and [-@Luo2017], and Magaloni [-@magaloni2010].

A related class of formal models holds that fraud operates as a distortion of the signal of the incumbent's true strength. In this type of model, such as those developed by Little [-@littleFraudMonitoringNoncompetitive2015] or Egorov and Sonin [-@Egorov2021], fraud may be used by incumbents to bluff their opponents into being more supportive than they would be if they knew the incumbent's true popularity. However, greater transparency makes such bluffs more costly. The typical result in these models is that protest risk can *partially* deter fraud. In equilibrium, the incumbent will generally choose the level of fraud anticipated by the opposition and no more\textemdash but also no less.

Fundamentally, both set of models share the assumption that protest is a centralized, strategic decision by opposition groups, and that fraud is a tool employed by weak (i.e. unpopular) incumbents. As a result, they predict that the highest risk of protest should obtain when incumbents use massive fraud but still produce a low incumbent vote-share, indicating that the incumbent's true popularity is very low. This, in turn, indicates to opposition actors that protest is more likely to succeed in unseating the incumbent. Incumbents in these models typically face a tradeoff between the short-term benefits of electoral manipulation for winning the election, and any weakness that relying on fraud reveals to the opposition [@Rozenas2016b]. As Beaulieu [-@Beaulieu2014a, p. 48] writes in this tradition, "As the opposition grows stronger, the amount of [manipulation] he will tolerate shrinks, and the incumbent who wishes to avoid protest will need to reduce the amount of manipulation she commits." These are normatively optimistic models upon which a sizable literature has been founded: fraud can be kept in check, and ultimately reduced, through the development of institutions that reveal information about election integrity: election monitoring [@Hyde2014;@daxecker2012], more independent courts [@chernykh2015third], media activity [@naiFourthEstate], and civil society coordination [@birch2017getting]. These institutions improve election integrity, it is claimed, by increasing the likelihood that fraud will be observed and thus, incumbent weakness exposed.

However, an alternative set of models have also been advanced. A second worldview argues that election manipulation can indicate a well-resourced, well-organized, powerful incumbent (independent of its popularity). Under this view, extensive election manipulation can be a show of strength, which induces cooperation with the incumbent from other societal actors. Since this view holds that more manipulation can further entrench the incumbent, it is a normatively pessimistic view.

Most persuasively articulated by Simpser [-@simpser2013], this view holds that election manipulation demonstrates incumbent strength. As he writes, an incumbent that demonstrates the ability to engage in widespread manipulation "is also likely to have the resources, capacities, and inclinations to overcome or punish opponents, reward supporters, and circumvent the law" [@simpser2013, p. 6]. A relatively popular incumbent ("strong" in the language used by the first set of models) may nonetheless manipulate heavily in order to produce a very large majority [@magaloni2006], to induce regime insiders to remain cooperative [@gehlbach2015electoral], or because of a bandwagon effect among front-line agents [@rundlett2016deliver]. By demonstrating otherwise difficult-to-observe capabilities, large-scale election manipulation can solve a coordination problem for the incumbent, securing the cooperation of elites [@simpser2013]. This can, in turn, fix expectations that the incumbent will continue to rule in the future, further enhancing its staying power [@hale_patronal_2014]. This argument is in keeping with Seeberg's [-@Seeberg2019] finding that higher-capacity states are associated with more election fraud.

If election manipulation can be an indicator of regime strength, it may serve to deter protest under some circumstances, rather than the other way around [@simpser2013]. This leaves protest, when it does occur in the presence of large-scale manipulation, as something of a puzzle if protest behavior is largely strategic. For this reason, the manipulation-as-strength view comports more easily with another established view of the causes of electoral protest. Researchers in this tradition have argued that electoral protest occurs largely unrelated to strategic concerns about the relative strength of the incumbent and the opposition. In these models, electoral protest occurs because manipulation galvanizes citizens who hold diverse grievances, because economic conditions are poor, or simply because the bare fact of holding an election creates a moment of opportunity for opposition organization irrespective of the level of manipulation [@Pop-Eleches2015]. The central feature of these models is that protest decisions are generally made without consideration for the likelihood that protests will overthrow the incumbent.^[To be clear, this is the narrow sense in which 'strategic' and 'non-strategic' is used here and in protest-deterrent models. Protest behavior may still be rational in a broader sense without being narrowly tied to the expected likelihood of ousting the government. For instance, protesters may believe it is unlikely they will be individually punished for participating in protests, making it individually rational for them to voice their grievances.]

@@

Rational protest entrepreneurs might seek to initiate protest to take advantage of this popular discontent for non-strategic reasons, in order to press for improved competitiveness in the *next* election or to bargain for policy concessions [@chernykh2014;@robertson_protest_2010]. They may stage protests in order to identify new participants, forge new social ties [@barberaModelProtestsRevolution2019] and test new tactics. They may seek to communicate information about the government's unpopularity, to set the stage for larger, future protests [@lohmannDynamicsInformationalCascades1994]. Protests may be staged as part of internal factional rivalries within the opposition [@Buckles2017], or to drive media coverage of issues important to the opposition [@huangMediaProtestDiffusion2019;@jenningsStreetDemonstrationsMedia2019]. Or, a vanguard of protesters may take to the streets simply because their distaste for the government outweighs any risk of sanction [@lohmannDynamicsInformationalCascades1994].  These and other motivations for initiating protest are theoretically significant because their rationality (for protesters) does not hinge on the likelihood of overturning the government, as classic models assume. As a result, it is possible that we may observe the initiation of anti-government electoral protest even when the likelihood of such protest overturning the election (or the government itself) is very low.

Empirical work has found some support for the microfoundations of this theory. For example, ballot-stuffing and voter coercion have been shown to generate stronger feelings of anger than pre-election manipulation of ballot access [@Szakonyi2021], and access to information about manipulation appears to reduce support for the ruling party [@Reuter2021], increase support for protest and reduce the perceived legitimacy of the regime [@Williamson2021]. This argument has found support in studies of the relationship between economic grievances, manipulated elections, and protest [@Rod2019;@Brancati2016].

Taking this all together, it becomes clear that if researchers focus only on protest occurrence as a dependent variable, they cannot distinguish between two possible data-generating processes: one in which manipulation signals weakness and electoral protest is a strategic response, and one in which electoral manipulation indicates regime strengths and protest occurs for a variety of reasons with limited connection to the odds of removing the government. In the former, protest is predicted to occur most often when manipulation is high and the incumbent's vote-share is low, since this provides a strong signal that the incumbent's true popularity is low. In the latter, protest is most likely in the same condition because the incumbent's low popularity indicates widespread grievances, which the high level of manipulation serves to coordinate despite the incumbent's organizational strengths. Which of these data-generating processes predominates is of significant theoretical importance of understanding authoritarianism and democratization, since the latter view suggests much dimmer prospects for democratization than the former.

@@

Since the empirical record is mixed, and two theories make similar predictions with regard to protest initiation, it is necessary to generate new hypotheses that can distinguish between the two models of election manipulation and protest that have been advanced. In particular, do post-election protest movements encounter weak incumbents as signaling models predict? If not, the deterrence model is undermined.

>>>>

# Literature review: two assumptions about election manipulation and mass protest

As sketched briefly above, a fruitful literature has examined the connections between election integrity and mass protest. Research in the regime-weakness model of manipulation typically concludes that as incumbents become increasingly fearful of street protest, they dial back on manipulation\textemdash a potentially pro-democracy outcome. These models hinge on two assumptions, as discussed below. First, election manipulation is conceptualized as a short-term tool used instrumentally to secure victory in the election [@lehoucq2002]; it communicates only that the incumbent could not count on winning the election cleanly. In such models incumbents are typically characterized as 'weak' or 'strong' based on their true support in the electorate: strong incumbents are those that are likely to defeat protest movements, while weak incumbents are those that are likely to be ousted.

This true level of support can be obscured prior to elections due to pervasive preference falsification [@Kuran1991] and the low level of political information common in authoritarian regimes [@guriev2020]. However, low popularity (i.e. weakness, in these models) can be revealed when incumbents need to resort to fraud in order to secure re-election. As a result, the opposition perceives election fraud as a reliable signal of incumbent weakness, and will stage protests in order to push out the weak incumbent. For example, the model developed by Little, Tucker, and LaGatta [-@Little2015] predicts that "Just as stronger election results for the incumbent lead to less protest because they indicate that she is popular, reports of more fraud lead to more protest because they indicate that the incumbent is, for a given election result, less popular" (p. 1153). In equilibrium, this can lead weaker incumbents to step down rather than face protest due to massive fraud\textemdash a pro-democracy outcome in which stronger opposition groups deter manipulation by incumbents. Similar results are developed by Chernykh and Svolik [-@chernykh2015third], Luo and Rozenas [-@Luo2020] and [-@Luo2017], and Magaloni [-@magaloni2010].

Some more recent models treat election manipulation as a distortion of the electoral signal of the incumbent's true strength, but are still located within the regime-weakness framework. In this type of model, such as those developed by Little [-@littleFraudMonitoringNoncompetitive2015] or Egorov and Sonin [-@Egorov2021], fraud may be used by incumbents to bluff their opponents into being more supportive than they would be if they knew the incumbent's true popularity. However, greater transparency makes such bluffs more costly, by making it easier for opposition actors to distinguish signals of true strength (i.e. genuine popularity) from noise (i.e. manufacured votes). The typical result in these models is that protest risk can *partially* deter fraud. In equilibrium, the incumbent will generally choose the level of fraud anticipated by the opposition and no more\textemdash but also no less. In sum, regime-weakness models of fraud typically hold that the level of election manipulation carries no signal of the incumbent's ability to defeat protesters; on the contrary, it only carries signals of weakness.

The second assumption in the regime-weakness framework concerns the nature of opposition protest. The framework is tightly linked to a view of electoral protest that is a centralized decision by rational actors pursuing narrowly strategic goals. As a result, they predict that the likelihood of protest should be highest when incumbents use massive fraud but still produce a low incumbent vote-share. This signal of weakness, in turn, indicates to opposition actors that protest is more likely to succeed in unseating the incumbent. Incumbents in these models typically face a tradeoff between the short-term benefits of electoral manipulation for winning the election, and any weakness that relying on fraud reveals to the opposition [@Rozenas2016b]. As Beaulieu [-@Beaulieu2014a, p. 48] writes in this tradition, "As the opposition grows stronger, the amount of [manipulation] he will tolerate shrinks, and the incumbent who wishes to avoid protest will need to reduce the amount of manipulation she commits."

These are normatively optimistic models upon which a sizable literature has been founded: manipulation can be kept in check, and ultimately reduced, through the development of institutions that reveal information about election integrity: election monitoring [@Hyde2014;@daxecker2012], more independent courts [@chernykh2015third], media activity [@naiFourthEstate], and civil society coordination [@birch2017getting]. These institutions improve election integrity, it is claimed, by increasing the likelihood that fraud will be observed and thus, incumbent weakness exposed.

However, there is also a body of research that challenges the two assumptions upon which the regime-weakness model is founded. The assumption on the nature of protest as centralized and narrowly strategic, in particular, has been challenged in the literature on electoral protest. Recent research in this tradition has argued that electoral protest occurs largely unrelated to strategic concerns about the relative strength of the incumbent and the opposition. In these models, electoral protest occurs because manipulation galvanizes citizens who hold diverse grievances [@tucker2007], because economic conditions are poor [@Brancati2016;@Beaulieu2014a], or simply because the bare fact of holding an election creates a moment of opportunity for opposition organization irrespective of the level of manipulation [@Pop-Eleches2015]. The central feature of these models is that protest decisions are generally made without consideration for the likelihood that protests will overthrow the incumbent.

@@

Rational protest entrepreneurs might seek to initiate protest to take advantage of this popular discontent for broadly strategic reasons. They may wish to pressure the incumbent in order to push for improved competitiveness in the *next* election or to bargain for policy concessions [@chernykh2014;@robertson_protest_2010]. They may stage protests in order to identify new participants, forge new social ties [@barberaModelProtestsRevolution2019] and test new tactics. They may seek to communicate information about the government's unpopularity, to set the stage for larger, future protests [@lohmannDynamicsInformationalCascades1994]. Protests may be staged as part of internal factional rivalries within the opposition [@Buckles2017], or to drive media coverage of issues important to the opposition [@huangMediaProtestDiffusion2019;@jenningsStreetDemonstrationsMedia2019]. Or, a vanguard of protesters may take to the streets simply because their distaste for the government outweighs any risk of sanction [@lohmannDynamicsInformationalCascades1994]. These and other motivations for initiating protest are theoretically significant because their rationality (for protesters) does not hinge on the likelihood of overturning the government, as regime-weakness models of manipulation assume. As a result, it is possible that we may observe the initiation of anti-government electoral protest even when the likelihood of such protest overturning the election (or the government itself) is very low.

Finally, an alternative theory of election manipulation challenges the assumption that an incumbent's ability manufacture votes communicates nothing about its ability to defeat a protest movement. This view, most persuasively articulated by Simpser [-@simpser2013], holds that increasing levels of election manipulation can indicate a well-resourced, well-organized, powerful incumbent (independent of its popularity); these resources and capacities, in turn, may help the incumbent weather a protest movement. Under this view, extensive election manipulation can be a show of strength, which induces cooperation with the incumbent from other societal actors and may serve to deter protest [@simpser2013]. Since this view holds that more manipulation can further entrench the incumbent, it is a normatively pessimistic one.

As Simpser writes, an incumbent that demonstrates the ability to engage in widespread manipulation "is also likely to have the resources, capacities, and inclinations to overcome or punish opponents, reward supporters, and circumvent the law" [@simpser2013, p. 6]. A relatively popular incumbent ("strong" in the language used by the first set of models) may nonetheless manipulate heavily in order to produce a very large majority [@magaloni2006], to induce regime insiders to remain cooperative [@gehlbach2015electoral], or because of a bandwagon effect among front-line agents [@rundlett2016deliver]. By demonstrating otherwise difficult-to-observe capabilities, large-scale election manipulation can solve a coordination problem for the incumbent, securing the cooperation of elites [@simpser2013]. This can, in turn, fix expectations that the incumbent will continue to rule in the future, further enhancing its staying power [@hale_patronal_2014]. This argument is in keeping with Seeberg's [-@Seeberg2019] finding that higher-capacity states are associated with more election fraud.

We can thus consider four possible states of the world, based on which of the foregoing assumptions are true in fact: one where both assumptions hold, one where neither holds, and two where only one assumption holds. The normatively optimistic view of the relationship between election manipulation and protest requires both assumptions to be true: electoral manipulation must communicate little information about incumbents' ability to outlast protest, and protest initiation must be narrowly strategic. If either one of these assumptions fails to hold, the deterrence logic is undermined. However, empirically distinguishing between these four possible states of the world is further complicated by the fact that work on this subject, whether empirical or purely formal, has largely utilized a binary measure of protest occurrence as the outcome of interest. Unfortunately, as Table 1 shows, three of four possible states of the world make the same prediction: that increasing manipulation will result in an increased likelihood of protest.

```{=latex}
% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[]
\begin{tabular}{@{}llll@{}}
\toprule
\multicolumn{1}{c}{\textbf{Protest type}} &
  \multicolumn{1}{c}{\textbf{\begin{tabular}[c]{@{}c@{}}Likelihood\\ of \\ Protest\end{tabular}}} &
  \multicolumn{1}{c}{\textbf{Mechanism}} &
  \textbf{Example sources} \\ \midrule
\multicolumn{1}{l|}{} &
  \multicolumn{3}{c}{Regime-weakness model} \\ \cmidrule(l){2-4} 
\multicolumn{1}{l|}{\begin{tabular}[c]{@{}l@{}}Narrowly \\ strategic\end{tabular}} &
  Increases &
  \begin{tabular}[c]{@{}l@{}}More manipulation indicates \\ lower true popularity, and thus \\ lower likelihood that \\ incumbent will survive protests.\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Luo and Rozenas (2020); \\ Chernykh and Svolik (2015); \\ Little, Tucker, and \\ LaGatta (2015)\end{tabular} \\
\multicolumn{1}{l|}{\begin{tabular}[c]{@{}l@{}}Broadly \\ strategic\end{tabular}} &
  Increases &
  \begin{tabular}[c]{@{}l@{}}More manipulation employed to \\ compensate for deeper \\ grievances; 'stolen' election as \\ trigger.\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Beaulieu (2014); \\ Brancati (2016); \\ Tucker (2007)\end{tabular} \\ \midrule
\multicolumn{1}{l|}{} &
  \multicolumn{3}{c}{Regime-strength model} \\ \cmidrule(l){2-4} 
\multicolumn{1}{l|}{\begin{tabular}[c]{@{}l@{}}Narrowly \\ strategic\end{tabular}} &
  Decreases &
  \begin{tabular}[c]{@{}l@{}}More manipulation indicates \\ capacities that increase \\ likelihood incumbent survives\\ protest, offsetting low popularity.\end{tabular} &
  Simpser (2013) \\
\multicolumn{1}{l|}{\begin{tabular}[c]{@{}l@{}}Broadly \\ strategic\end{tabular}} &
  Increases &
  \begin{tabular}[c]{@{}l@{}}More manipulation employed to\\ compensate for deeper \\ grievances; 'stolen' election as\\ trigger.\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Beaulieu (2014); \\ Brancati (2016);\\ Tucker (2007)\end{tabular}
\end{tabular}
\caption{Table 1: Predictions for protest initiation under the regime-weakness model of electoral manipulation}
\label{tab:protest-expectation}
\end{table}
```

@@

Nevertheless, there are considerable theoretical and substantive stakes in knowing which of these four possible data-generating processes is most likely; in only one does increasing protest risk actually improve electoral integrity. Since the empirical record is mixed, and three of the four possible models make similar predictions with regard to protest initiation, it is necessary to generate new hypotheses.  In particular, the remainder of this manuscript will attempt to distinguish between regime-strength and regime-weakness models of manipulation by developing hypotheses about protest *duration* under each framework. Specifically, I will argue that protest duration helps reveal the true relative balance of power between incumbents and opposition actors.

<!-- NOT SURE WHERE THIS GOES:
Empirical work has found some support for the microfoundations of this theory. For example, ballot-stuffing and voter coercion have been shown to generate stronger feelings of anger than pre-election manipulation of ballot access [@Szakonyi2021], and access to information about manipulation appears to reduce support for the ruling party [@Reuter2021], increase support for protest and reduce the perceived legitimacy of the regime [@Williamson2021]. This argument has found support in studies of the relationship between economic grievances, manipulated elections, and protest [@Rod2019;@Brancati2016].
-->
